---
title: "MAE building suggestion"
author: "Lisa Arnalot"
date: last-modified
date-format: "[Last Updated on] MMMM DD, YYYY"
format: 
  html:
    code-fold: true
    toc: true
    toc-depth: 5
    toc-location: left
    theme: paper
    embed-resources: true
execute:
  cache: refresh
project:
  execute-dir: project
editor: 
  markdown: 
    wrap: 80
number-sections: true
---

# MultiAssayExperiment

## Introduction

In the FRESH project, data from multi-omics experiments were generated across
different biological compartments and timepoints, with each assay producing
matrices of different dimensions. A flat data structure cannot accommodate this
heterogeneity without duplication or loss of linkage. We therefore adopt the
`MultiAssayExperiment` (MAE) framework proposed by Ramos et al. (2017), which
was designed precisely for this use case.

The MAE enforces a strict separation between three layers of information:

**ExperimentList** holds the assay data themselves — one entry per omics
modality (e.g. `$16S`, `$metagenomics`, `$proteomics`). Each entry is a
`SummarizedExperiment` or matrix whose columns correspond to physical samples.
At this stage the list is empty; assays will be added progressively as
sequencing and analysis outputs become available.

**colData** is the primary sample registry — one row per biological unit, here
defined as one participant (`pid`) at one sampling occasion
(`visit_code × study_day`). It is the anchor to which all assay data are
ultimately linked. The row names of `colData` serve as the primary keys of the
entire structure.

**sampleMap** is a three-column linking table (`assay`, `primary`, `colname`)
that maps each column of each assay matrix back to exactly one row of `colData`.
It cannot be populated until assay matrices are available, because the `assay`
column must contain names present in the `ExperimentList`, and the `colname`
column must match the actual column names used in each matrix. The LIMS manifest
is stored in `metadata` to make this mapping possible once data arrive.

**metadata** is a free-form named list used to store supplementary information
that does not fit the row-per-sample structure of `colData`. Here it holds the
raw LIMS manifest, global date anchors, CRF data frames, reference identifier
tables, and a placeholder for derived analytical variables.

::: callout-note
*Are there additional data sources that should be incorporated into this MAE at
this stage or registered as future assays?* There are several warnings across
the document, we can delete them if you would like now that the code runs.
:::

## Set up

### Loading libraries and data

```{r}
#| warning: false
library("MultiAssayExperiment")
library(stringr)
library(readxl)
library(dplyr)
library(lubridate)
library(purrr)
```

```{r}
tmp <- fs::dir_map("R", source)
```

Pre-cleaned data objects produced by upstream cleaning scripts (fresh_surveys
repository) are loaded from `derived_data/Clean/`.

```{r}
files <- list.files(
  path       = str_c(get_data_dir(), "derived_data/Clean/"),
  pattern    = "\\.rds$",
  full.names = TRUE
)

for (f in files) {
  name <- tools::file_path_sans_ext(basename(f))
  assign(name, readRDS(f), envir = .GlobalEnv)
}

```

The LIMS manifest is loaded, it is a record of all physical samples collected in
the FRESH project.

```{r}
manifest <- read_xlsx(
  str_c(get_data_dir(), "HPP_LIMS/FRESH HPP127 Samples-Comprehensive-01 October 2025.xlsx")
)|> as.data.frame()
```

```{r}
survey_raw <- readRDS(str_c(get_data_dir(),"REDCap exports/raw_export_20240730/study_surveys.rds"))

acutes = read_excel(str_c(get_data_dir(), "Acutes tables/_2024.9_FRESH Acute data_DR 22 Feb 2025.xlsx"), sheet = "Acute List (all)") %>% rename(study_id = `PID...2`) %>% 
  mutate(study_id = factor(str_remove_all(study_id, "-")))
```

::: callout-important
*Are there any more data available so far that could be included at this stage ?
Should there be some kind of cleaning there for the acutes for instance ?* add
the cohort/enrollment group (from where ?) is it : clean_blood_mucosal$group
& survey_raw$group how can we handle the NA ?
:::

### Harmonising participant identifiers

#### Reference crosswalk

Participant identifiers were recorded inconsistently across data collection
instruments. The canonical identifier labelled `pid` is `clean_study_id` from
`cleaned_visit_codes_and_dates`, which was established during the visit-code
cleaning pipeline. A crosswalk table maps every raw identifier variant to its
canonical form:

```{r}
id_crosswalk <- cleaned_visit_codes_and_dates |>
  select(raw_study_id, clean_study_id) |>
  distinct()

# checking for duplicates
stopifnot(
  "Duplicate raw_study_id found in crosswalk" =
    id_crosswalk |> count(raw_study_id) |> filter(n > 1) |> nrow() == 0
)

# flagging known issues from cleaning-visit_codes.Rmd
known_unresolvable_pids <- c(
  # canonical formatted versions (appear in validation dfs)
  "127-33-2455-2639",  # confirmed typo, deleted from visit cleaning pipeline
  "127-33-2708-2833",  # confirmed typo, deleted from visit cleaning pipeline
  "127-33-4-17-0000",  # malformed ID, non-standard format
  "127-33- 3831-0000", # malformed ID, contains whitespace
  # raw unformatted versions (appear in baseline dfs — same participants,
  # absent from crosswalk so harmonise_pid could not convert them)
  "1273324552639",     # raw form of 127-33-2455-2639
  "1273327082833"      # raw form of 127-33-2708-2833
)
```

#### Harmonisation of pid

The function below looks for a `study_id` column in any data frame, joins it to
the crosswalk, and replaces it with the canonical `pid`. If a `pid` column
already exists, it is validated against the crosswalk. In both cases, unmatched
values fall back to their raw form and trigger a warning so that new
discrepancies are flagged.

```{r}
harmonise_pid <- function(df) {
  if ("study_id" %in% names(df)) {
    df <- df |>
      mutate(study_id = as.character(study_id)) |>
      left_join(id_crosswalk, by = c("study_id" = "raw_study_id")) |>
      mutate(pid = coalesce(clean_study_id, study_id)) |>
      select(-study_id, -clean_study_id)
  } else if ("pid" %in% names(df)) {
    # verify existing pid matches canonical clean_study_id
    unmatched <- unique(
      df$pid[!(df$pid %in% id_crosswalk$clean_study_id)]
    )
    if (length(unmatched) > 0) {
      warning(
        "harmonise_pid: pid column exists but ",
        length(unmatched),
        " value(s) not found in crosswalk — check these:\n  ",
        paste(unmatched, collapse = "\n  "),
        call. = FALSE
      )
    }
  }
  df
}

clean_3mo_risk <- clean_3mo_risk |>
  select(-uid)

clean_baseline_2                               <- harmonise_pid(clean_baseline_2)
clean_baseline_intake                          <- harmonise_pid(clean_baseline_intake)
clean_baseline_intake_with_subjective_cleaning <- harmonise_pid(clean_baseline_intake_with_subjective_cleaning)
clean_blood_mucosal                            <- harmonise_pid(clean_blood_mucosal)
clean_blood_mucosal_with_subjective_cleaning   <- harmonise_pid(clean_blood_mucosal_with_subjective_cleaning)
validation_baseline_intake                     <- harmonise_pid(validation_baseline_intake)
validation_baseline_2                          <- harmonise_pid(validation_baseline_2)
clean_3mo_risk                                 <- harmonise_pid(clean_3mo_risk)
```

Unmatched identifiers are then classified as either known (excluded by design)
or unexpected (requiring investigation before proceeding).

```{r}
check_pid_coverage <- function(df, df_name) {
  unmatched <- unique(
    df$pid[!(df$pid %in% cleaned_visit_codes_and_dates$clean_study_id)]
  )
  known   <- intersect(unmatched, known_unresolvable_pids)
  unknown <- setdiff(unmatched, known_unresolvable_pids)

  if (length(known) > 0) {
    message(
      df_name, ": ", length(known),
      " known unresolvable pid excluded by design from visit cleaning:\n  ",
      paste(known, collapse = "\n  ")
    )
  }
  if (length(unknown) > 0) {
    warning(
      df_name, ": ", length(unknown),
      " UNEXPECTED unmatched pid investigate before proceeding:\n  ",
      paste(unknown, collapse = "\n  "),
      call. = FALSE
    )
  }
  invisible(list(known = known, unknown = unknown))
}

unresolved_ids <- list(
  clean_baseline_2                               = check_pid_coverage(clean_baseline_2,                               "clean_baseline_2"),
  clean_baseline_intake                          = check_pid_coverage(clean_baseline_intake,                          "clean_baseline_intake"),
  clean_baseline_intake_with_subjective_cleaning = check_pid_coverage(clean_baseline_intake_with_subjective_cleaning, "clean_baseline_intake_with_subjective_cleaning"),
  clean_blood_mucosal                            = check_pid_coverage(clean_blood_mucosal,                            "clean_blood_mucosal"),
  clean_blood_mucosal_with_subjective_cleaning   = check_pid_coverage(clean_blood_mucosal_with_subjective_cleaning,   "clean_blood_mucosal_with_subjective_cleaning"),
  validation_baseline_intake                     = check_pid_coverage(validation_baseline_intake,                     "validation_baseline_intake"),
  validation_baseline_2                          = check_pid_coverage(validation_baseline_2,                          "validation_baseline_2"),
  clean_3mo_risk                          = check_pid_coverage(clean_3mo_risk,                          "clean_3mo_risk")
)

```

## Building **colData** based on LIMS manifest

The LIMS manifest is the only source that records every physical sample
collected, regardless of whether downstream assay data exist yet. Building
`colData` from it ensures that every sampling occasion is represented in the MAE
even before sequencing outputs are available. Rows in `colData` that have no
matching assay data do not appear in assay-specific subsets.

### Cleaning and parsing the manifest

Several data quality issues were identified in the raw manifest and are fixed
here.

The volume value contains a mix of units (`ml`, `ML`, `ul`, `UL`), European
decimal notation (`1,0`), trailing punctuation (`1.0.`), and the string `EMPTY`.
All values are normalised to microlitres. Bare integers below 5 are assumed to
have been entered in millilitres and are multiplied by 1000 accordingly. The raw
value is retained as `volume_raw`.

::: callout-warning
Do you agree with the fixes done here ? is it important at this stage ?
:::

```{r}
parse_volume_ul <- function(x) {
  x_clean <- str_trim(str_to_upper(x)) |>
    str_replace_all(",", ".") |> 
    str_replace_all("(?<=\\d)\\.\\s+(?=\\d)", ".") |> 
    str_replace_all("(?<=\\d)\\s+(?=[A-Z])", "") |>
    str_replace_all("(?<=[A-Z])\\s+(?=\\d)", "") |>
    str_replace_all("\\.$", "") |>
    str_replace_all("(?<=\\d)M$", "ML")
  result <- case_when(
    x_clean == "EMPTY"                          ~ 0,
    str_detect(x_clean, "^[0-9.]+ML$")         ~ as.numeric(str_extract(x_clean, "[0-9.]+")) * 1000,
    str_detect(x_clean, "^[0-9.]+UL$")         ~ as.numeric(str_extract(x_clean, "[0-9.]+")),
    str_detect(x_clean, "^[0-9.]+$")           ~ {
      val <- as.numeric(x_clean)
      if_else(val < 5, val * 1000, val)
    },
    TRUE~ NA_real_
  )
  unrecognised <- x[is.na(result) & !is.na(x)]
  if (length(unrecognised) > 0) {
    warning(
      "parse_volume_ul: ", length(unrecognised),
      " unrecognised volume pattern(s) → NA:\n  ",
      paste(unique(unrecognised), collapse = "\n  "),
      call. = FALSE
    )
  }
  result
}

```

```{r}
manifest_clean <- manifest |>
  rename(
    pid         = `\r\nPID`,
    sample_date = `\r\nSample_Date`,
    visit_code  = `\r\nTimepoint`,
    volume_raw  = `\r\nValue`,       # keep raw 
    sample_type = `Sample Type`,
    n_vials     = `No of vials`,
    storage     = `Storage`,
    lims_number = `LIMS_Number`
  ) |>
  select(pid, sample_date, visit_code, volume_raw,
         sample_type, n_vials, storage, lims_number) |>
  mutate(
    sample_date = as_date(sample_date),
    volume_ul   = parse_volume_ul(volume_raw), # use the function to fix the raw volume
    lims_number = as.character(lims_number)
  ) |>
  filter(!is.na(pid), !is.na(lims_number))

# Audit: flag any values that could not be fixed
unparseable <- manifest_clean |>
  filter(is.na(volume_ul), !is.na(volume_raw)) |>
  count(volume_raw)

if (nrow(unparseable) > 0) {
  warning(
    "parse_volume_ul: ", nrow(unparseable),
    " unrecognised volume pattern(s) → NA:\n  ",
    paste(unparseable$volume_raw, collapse = "\n  "),
    call. = FALSE
  )
} else {
  message("parse_volume_ul: all values parsed successfully")
}

```

#### Computing study_day

Rather than relying on nominal visit dates alone for temporal ordering, we
compute `study_day` as the number of days since each participant's earliest
sample in the manifest. This anchor is computed once from the full manifest so
that the value is consistent across `colData` and the future `sampleMap`:

```{r}
global_anchors <- manifest_clean |>
  group_by(pid) |>
  summarise(anchor_date = min(sample_date, na.rm = TRUE), .groups = "drop")
```

#### Constructing the primary key

The primary key uniquely identifies each row of `colData`. We use
`pid_visitcode_studyday` rather than `pid` alone (not unique across visits) or
`pid_visitcode` alone (32 visit codes have samples on two distinct dates).
Including `study_day` resolves this ambiguity while keeping the key
human-readable. The three rows with a missing `visit_code` use `pid_studyday` as
a fallback.

```{r}
col_data_df <- manifest_clean |>
  left_join(global_anchors, by = "pid") |>
  mutate(study_day = as.integer(sample_date - anchor_date)) |>
  select(-anchor_date) |>
  group_by(pid, visit_code, sample_date, study_day) |>
  summarise(.groups = "drop") |>
  mutate(
    primary_key = case_when(
      !is.na(visit_code) ~ paste(pid, visit_code, study_day, sep = "_"),
      TRUE               ~ paste(pid, study_day, sep = "_")
    )
  ) |>
  tibble::column_to_rownames("primary_key")

stopifnot(
  "primary_key is not unique in colData" =
    n_distinct(rownames(col_data_df)) == nrow(col_data_df)
)

message(
  "colData: ", nrow(col_data_df), " rows across ",
  n_distinct(col_data_df$pid), " participants"
)
```

::: callout-warning
**32 visit codes have samples on two distinct dates (due to split collections or
data entry issues?)** should we investigate to fix this ? Right now, they are
retained as separate `colData` rows (one per date) rather than collapsed.
:::

manifest_clean \|\> group_by(pid, visit_code) \|\> mutate(n_dates =
n_distinct(sample_date)) \|\> filter(n_dates \> 1) \|\> arrange(pid, visit_code,
sample_date) \|\> select(pid, visit_code, sample_date, sample_type, lims_number,
n_vials) \|\> print(n = Inf)

#### Adding visit codes to ColData

Where available, visit codes from the cleaned visit-code pipeline are used to
fill or confirm the `visit_code` field. This handles cases where the LIMS
recorded a slightly different code from the one validated during clinical data
cleaning.

```{r}
visit_code_lookup <- cleaned_visit_codes_and_dates |>
  filter(!clean_study_id %in% known_unresolvable_pids) |>
  select(
    pid        = clean_study_id,
    visit_date = clean_visit_date,
    clean_visit_code
  ) |>
  distinct() |>
  group_by(pid, visit_date) |>
  slice(1) |>        # keep one row per pid x visit_date
  ungroup()

col_data_df <- col_data_df |>
  as.data.frame() |>
  mutate(rn = rownames(col_data_df)) |>
  left_join(visit_code_lookup, by = c("pid", "sample_date" = "visit_date")) |>
  mutate(visit_code = coalesce(clean_visit_code, visit_code)) |>
  select(-clean_visit_code) |>
  tibble::column_to_rownames("rn")

stopifnot(
  "row names not unique after visit code enrichment" =
    !any(duplicated(rownames(col_data_df)))
)

col_data_df <- DataFrame(col_data_df)

message(
  "colData after visit code enrichment: ", nrow(col_data_df), " rows across ",
  n_distinct(col_data_df$pid), " participants"
)
```

#### Adding LIMS number to ColData

```{r}

lims_per_occasion <- manifest_clean |>
  left_join(global_anchors, by = "pid") |>
  mutate(
    study_day   = as.integer(sample_date - anchor_date),
    primary_key = case_when(
      !is.na(visit_code) ~ paste(pid, visit_code, study_day, sep = "_"),
      TRUE               ~ paste(pid, study_day, sep = "_")
    )
  ) |>
  group_by(primary_key) |>
  summarise(
    lims_numbers  = list(unique(lims_number)),
    sample_types  = list(unique(sample_type)),
    n_vials_total = n(),
    .groups = "drop"
  )

col_data_df <- col_data_df |>
  as.data.frame() |>
  tibble::rownames_to_column("primary_key") |>
  left_join(lims_per_occasion, by = "primary_key") |>
  tibble::column_to_rownames("primary_key") |>
  mutate(
    lims_numbers = I(lims_numbers),   # I() protects list-columns from being
    sample_types = I(sample_types)    # silently dropped by DataFrame()
  ) |>
  DataFrame()

stopifnot(
  "LIMS join duplicated colData rows" =
    nrow(col_data_df) == n_distinct(rownames(col_data_df)),
  "lims_numbers missing from colData" =
    "lims_numbers" %in% names(col_data_df),
  "DataFrame conversion dropped rows" =
    nrow(col_data_df) > 0
)

message(
  "colData enriched with LIMS inventory: ",
  nrow(col_data_df), " rows, ",
  sum(lengths(col_data_df$lims_numbers)), " total vials registered"
)
```

## Building the **SampleMap**

The `sampleMap` requires three pieces of information: the assay name, the
`colData` primary key, and the column name used in the assay matrix. The first
two can be derived from the manifest, but the third — the column identifier
actually used in each sequencing output — is not yet available. Critically,
`sample_type` in the LIMS (e.g. `CVL_PELLET`) describes physical material, not
an analytical assay: the same vial type could yield 16S amplicon data,
metagenomics, and proteomics data depending on which analyses were run. The
`sampleMap` will therefore be populated assay by assay as data arrive. The
manifest is stored in `metadata` to make that future join straightforward. The
primary keys are nonetheless pre-computed on the manifest so that the linkage
logic is already in place.

```{r}
manifest_clean <- manifest_clean |>
  left_join(global_anchors, by = "pid") |>
  mutate(study_day = as.integer(sample_date - anchor_date)) |>
  select(-anchor_date)

# verify
message(
  "study_day NAs in manifest_clean: ",
  sum(is.na(manifest_clean$study_day))
)

sample_map_df <- manifest_clean |>
  mutate(
    primary_key = case_when(
      !is.na(visit_code) ~ paste(pid, visit_code, study_day, sep = "_"),
      TRUE               ~ paste(pid, study_day, sep = "_")
    )
  ) |>
  filter(primary_key %in% rownames(col_data_df)) |>
  transmute(
    assay   = sample_type, # placeholder : it will be recoded per assay
    primary = primary_key,
    colname = lims_number # placeholder : it will be recoded per matrix colnames
  ) |>
  distinct() |>
  DataFrame()
```

::: callout-important
**sampleMap is a placeholder** The `assay` column currently contains LIMS sample
types, not analytical assay names. The `colname` column contains LIMS numbers,
which may or may not match the column names used in assay outputs. Both must be
reconciled when each assay matrix is added.
:::

## Checking CRF coverage for future **metadata**

Before assembling the MAE, we verify what proportion of participants in each CRF
can be linked to a `colData` row. Participants present in a CRF but absent from
`colData` would have no sample record in the LIMS and cannot be included in
assay-linked analyses.

```{r}
# check pid coverage in each CRF against colData
crf_pid_coverage <- list(
  baseline_intake              = clean_baseline_intake,
  baseline_intake_clean        = clean_baseline_intake_with_subjective_cleaning,
  baseline_intake_validation   = validation_baseline_intake,
  baseline_2                   = clean_baseline_2,
  baseline_2_validation        = validation_baseline_2,
  blood_mucosal                = clean_blood_mucosal,
  blood_mucosal_clean          = clean_blood_mucosal_with_subjective_cleaning,
  risk_3mo                     = clean_3mo_risk
) |>
  purrr::imap(function(df, name) {
    n_total    <- n_distinct(df$pid)
    n_matched  <- n_distinct(df$pid[df$pid %in% col_data_df$pid])
    n_unmatched <- n_total - n_matched
    message(
      name, ": ", n_matched, "/", n_total, " pids in colData",
      if (n_unmatched > 0) paste0(" (", n_unmatched, " unmatched)") else "matched"
    )
  })
```

## Assembling the skeleton MAE

At this stage, the MAE is assembled with an empty `ExperimentList`. Since the
`ExperimentList` is empty, any non-empty `sampleMap` would fail validation
immediately. The pre-computed version is stored in `metadata` for use when
assays are added (it would not work at this stage), purely as a reference table.
That will make building each assay's `sampleMap` slice a simple filter operation
rather than rebuilding the logic from scratch. All supporting data are stored in
`metadata` under a structured hierarchy so they remain accessible and versioned
alongside the MAE object itself.

```{r}
mae <- MultiAssayExperiment(
  experiments = ExperimentList(), #empty but will be completed with received data
  sampleMap   = DataFrame(  #empty but will be completed with received data
    assay   = factor(character()),
    primary = character(),
    colname = character()
  ),
  metadata = list(
    mae_creation_date = Sys.time(),
    sample_manifest   = list(
      source_file    = "FRESH HPP127 Samples-Comprehensive-01 October 2025.xlsx",
      raw            = manifest_clean,
      global_anchors = global_anchors, #now it is pid and and real date, but once the cleaning step will be done before - it should disappear in final MAE
      sample_map     = sample_map_df # placeholder : may be used for sampleMap building - it should disappear in final MAE
    ),
    crfs = list( #add dictionnary
      baseline_intake            = clean_baseline_intake,
      baseline_intake_clean      = clean_baseline_intake_with_subjective_cleaning,
      baseline_intake_validation = validation_baseline_intake,
      baseline_2                 = clean_baseline_2,
      baseline_2_validation      = validation_baseline_2,
      blood_mucosal              = clean_blood_mucosal,
      blood_mucosal_clean        = clean_blood_mucosal_with_subjective_cleaning,
      risk_3mo                   = clean_3mo_risk,
      survey_raw                 = survey_raw,
      acutes                     = acutes
    ),
    reference_id_vars = list( # I think this should be deleted from the mae if cleaning occured before
      crosswalk               = id_crosswalk, #the raw and cleaned study id - should not be kept in the mae
      visit_codes_and_dates   = cleaned_visit_codes_and_dates, #the cleaned/raw visit codes and dates - should not be kept in the mae
      unresolved_ids          = unresolved_ids, # list the unresolved pid issues - if cleaning occurs upstream - should not be kept in the mae
      known_unresolvable_pids = known_unresolvable_pids # among the unresolved pid issues the ones that were flagged in previous codes - should not be kept in the mae
    )
  )
)

# Assign colData after construction to bypass harmonisation against empty sampleMap
slot(mae, "colData") <- col_data_df

stopifnot(
  "colData not assigned correctly" =
    nrow(colData(mae)) == nrow(col_data_df),
  "colData rownames lost" =
    !is.null(rownames(colData(mae))),
  "metadata slots missing" =
    all(c("mae_creation_date", "sample_manifest",
          "crfs", "reference_id_vars")
        %in% names(metadata(mae)))
)

message(
  "MAE skeleton assembled successfully",
  "\n  colData    : ", nrow(colData(mae)), " rows across ",
                       n_distinct(colData(mae)$pid), " participants",
  "\n  LIMS vials : ", sum(lengths(colData(mae)$lims_numbers)),
                       " registered across colData rows",
  "\n  assays     : ", length(experiments(mae)),
                       " (empty — to be populated as data arrive)",
  "\n  sampleMap  : empty constructor; pre-computed linkage in ",
                       "metadata$sample_manifest$sample_map (",
                       nrow(metadata(mae)$sample_manifest$sample_map), " rows)",
  "\n  metadata   : ", length(metadata(mae)), " slots: ",
                       paste(names(metadata(mae)), collapse = ", ")
)
```

```{r}
#saveRDS(mae, str_c(get_data_dir(), "derived_data/mae_skeleton.rds"))
```

# Notes and remarks

## On replicates

DRAFT EXEMPLE adday primary colnames

I wondered how to handled duplicated; so taking an exemple of one sample run
twice for 16S.

This is exactly what the `colname` + `sampleMap` design handles. If LIMS number
`396236` was sequenced twice, producing two columns in the 16S matrix
(`396236_run1` and `396236_run2`), the `sampleMap` would have two rows with the
same primary but different colname values:

```         
assay   primary                      colname
16S     127-33-0001-003_1-011-0_0    396236_run1
16S     127-33-0001-003_1-011-0_0    396236_run2
```

This is explicitly supported by the MAE spec : duplicate primary values
(here`pid`x`visit_code`x`visit_day` ?) in the `sampleMap` represent replicates.
Both columns appear in the assay matrix, both point to the same `colData` row.
When subset by participant or visit, we get both replicates, and we keep all
information in the MAE until we know how to handle them (average, take best
quality, run sensitivity analysis).

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_following Laura and Lisa's
meeting\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

One participant has several visits for which different sample types were done.
These samples may have been aliquoted

```         
                                  ↗ Aliquot 1
                 ↗ specimen 1  → Aliquot 2                  ↗ Pool 1
                                  ↘ Aliquot 3 → DNAextract1 → Pool 2
                                              ↘ DNAextract2 → Pool 1
pid1 →  visit 1  →  specimen 2

                 ↘ specimen 3


                                ↗ Aliquot 1
                 ↗ specimen 1  → Aliquot 2               ↗ Pool 1
                               ↘ Aliquot 3 → DNAextract1 → Pool 2
                                            ↘ DNAextract2 → Pool 1
pid1 →  visit 2  →  specimen 2

                 ↘ specimen 3


pid1 -> visit 3
```

This would result in a table like this on :

```         
participant - visit - specimen - lims - aliquot - extraction - seq - assay   

pid1            1       1       XX1       1         1           1   amplicon_16S
pid1            1       1       XX1       2         1           1   amplicon_16S
pid1            1       1       XX1       3         1           1   amplicon_16S
pid1            1       1       XX1       3         1           2   amplicon_16S
pid1            1       1       XX1       3         2           1   amplicon_16S
pid1            1       2       XX2       1                         metagenomics
pid1            1       3       XX3       1                         cytokines
pid1            2       1       X21       1         1           1   amplicon_16S
pid1            2       1       X21       2         1           1   amplicon_16S
pid1            2       1       X21       3         1           1   amplicon_16S
pid1            2       1       X21       3         1           2   amplicon_16S
pid1            2       1       X21       3         2           1   amplicon_16S
pid1            2       2       X22       1                         metagenomics
pid1            2       3       X23       1                         cytokines
```

The samples we may have example for 16S amplicon:

```         
participant - visit - specimen - lims - aliquot - extraction - seq - colname
  1             1       SWAB     x41      1         1           1     SRR001
  1             1       SWAB     x41      1         2           1     SRR002
  1             1       SWAB     x41      1         2           2     SRR003
  
```

These informations should be stored in the metadata, along with the 'raw' sample
manifest.

```         
metadata(mae)$sample_manifest$sample_sheet_16S <- data.frame(
  pid = "127-33-0001-003", 
  visit = 1, 
  specimen = "SWAB", 
  lims = "x41", 
  aliquot = c(1, 1, 1), 
  extraction = c(1, 2, 2), 
  seq_run = c(1, 1, 2), 
  colname = c("SRR001","SRR002", "SRR003")
  )
```

Colname is the sequencing run identifier provided by the sequencing facility.

```         
assay     primary                     colname      
16S       127-33-0001-003_1-011-0_0   SRR001      
16S       127-33-0001-004_1-011-0_0   SRR002 ... 
```

## On LIMS numbers

I wondered how they will be handled, it depends on the assays data structure I
guess and hopefully it willlook like the following :

```         
colData row  →  sampleMap$colname  →  assay matrix column
                     ↑
               = lims_number
```

For each assay data received, verify the column names to match the
`lims_number`.

## Next steps : adding assays as data arrive

When a new assay matrix becomes available (e.g. 16S feature table, metagenomic
count matrix), the procedure is as follows.

**1. Identify which LIMS numbers correspond to this assay.**

Join the internal barcode or sample name to a LIMS number against
`metadata(mae)$sample_manifest$raw` to recover the `primary` key:

```{r}
#| eval: false
# Example: 16S assay arrives with columns named by LIMS number
assay_colnames  
  filter(lims_number %in% assay_colnames) |>
  mutate(
    primary = case_when(
      !is.na(visit_code) ~ paste(pid, visit_code, study_day, sep = "_"),
      TRUE               ~ paste(pid, study_day, sep = "_")
    )
  ) |>
  select(primary, colname = lims_number) |>
  distinct()
```

**2. Build the assay-specific sampleMap :**

```{r}
#| eval: false
samplemap_16s <- DataFrame(
  assay   = "16S",
  primary = assay_lims_map$primary,
  colname = assay_lims_map$colname
)
```

**3. Wrap the matrix in a SummarizedExperiment and add to the MAE:**

```{r}
#| eval: false
se_16s <- SummarizedExperiment(assays = list(counts = counts_16s))

mae <- c(
  mae,
  `16S` = se_16s,
  sampleMap = samplemap_16s
)

upsetSamples(mae)   # visualise sample coverage across assays
```

::: callout-note
**Assay naming convention** Assay names become permanent keys in the
`ExperimentList` and must match the `assay` column of every `sampleMap` slice
added thereafter.
:::

::: callout-important
**To do before the first assay is added** - Confirm that the column names in
each sequencing output are LIMS numbers, or obtain the crosswalk from the
sequencing barcodes. - Resolve the 32 pid × visit_code pairs with two sample
dates with the clinical team —\> determine whether these are split collections
(keep both rows) or entry errors (merge to one row). - Agree on canonical assay
names for `ExperimentList`to match in `sampleMap`.
:::

# Reference

Ramos, Marcel, Lucas Schiffer, Angela Re, et al. 2017. “Software for the
Integration of Multiomics Experiments in Bioconductor.” Cancer Research 77 (21):
e39–42. https://doi.org/10.1158/0008-5472.CAN-17-0344.
