---
title: "MAE building suggestion"
author: "Lisa Arnalot"
date: last-modified
date-format: "[Last Updated on] MMMM DD, YYYY"
format: 
  html:
    code-fold: true
    toc: true
    toc-depth: 5
    toc-location: left
    theme: paper
    embed-resources: true
execute:
  cache: true
project:
  execute-dir: project
editor: 
  markdown: 
    wrap: 80
number-sections: true
---

## MultiAssayExperiment introduction

In FRESH project, data from multi-omics experiments were generated, with
differet assays having different number of rows and columns. Hence for data
structure it is recommended to use a `MultiAssayExperiment` as proposed by Ramos
et al (2017), hereby referred as MAE.

The MAE include the followings :

- **ExperimentList** : a list including each assay data such as \$16S
  \$metagenomics \$...

- **colData** : one set of “primary” metadata that describes the ‘biological
  unit’; in FRESH it was decided that it would be one row per sample. Therefore,
  the **colData** should include : id_var \| uid \| pid \| visit_code \|
  rel_date \| ...

- **sampleMap** : a dataframe relating the **colData** to the experimental
  assays. One and only one row in **colData** matches the experimental
  observations. A row of **colData** can be associated with multiple
  experimental observations or no observations at all. It includes three
  columns - assay : the name of the different assay present in
  **ExperimentList** - primary (*pid?*): the information matching the
  **colData**. Values in this column must be present in the rownames of
  **colData**. - colname (*barcode?*): the sample names used in each assay. For
  each assay, all column names must be found in this column. Duplicate values
  are allowed, to represent replicates with the same overall experiment-level
  annotation.

- **metadata** : can be of any class and is used to store *named lists ?* of
  additional data, such as crfs...

  **drops** : a flag in case of mismatch. *It should not be in the final MAE ?*

## Loading data and libraries

```{r}
#| echo: false
if (!require("BiocManager"))
    install.packages("BiocManager")
BiocManager::install("MultiAssayExperiment")
```

```{r}
library("MultiAssayExperiment")
```

TBC...

# Reference

Ramos, Marcel, Lucas Schiffer, Angela Re, et al. 2017. “Software for the
Integration of Multiomics Experiments in Bioconductor.” Cancer Research 77 (21):
e39–42. https://doi.org/10.1158/0008-5472.CAN-17-0344.
