---
title: "Cox Model Demo"
execute:
    message: false
    warning: false
---

Load libraries and source necessary files.

```{r setup}
library(MultiAssayExperiment)
library(survival)
library(tidyverse)
library(yaml)
library(here)
library(fs)

tmp <- dir_map(here("src", "data"),   source, type = "file")
tmp <- dir_map(here("src", "models"), source, type = "file")
```

Define the event-oriented MAE and show the life table that will be used as the
response in the cox model.

```{r load-data}
emae <- from_mae(readRDS(here("data", "example_mae.rds")))
lt <- life_table(emae, cut_strategy = "visits")
emae
```

Read a configuration file. This is just a YAML with arguments for the particular
model we want to run.

```{r spec}
spec <- read_spec(here("conf", "models", "coxph_baseline.yml"))
spec
```

We can look at the fitted model. The `engine` output is the usual coxph output,
which can be summarized and plotted as usual.

```{r fit}
fit <- fit_engine(emae, spec)
fit
```

We can make predictions over new trajectories. We'll define the trajectory over
all timepoints seen in the training data (`fit$cuts`). The survival predictions
are made for each of those time intervals.

```{r newdata}
subjects <- colData(emae$mae) |>
    as_tibble() |>
    select(study_id, age_at_start) |>
    slice(1:3)

trajectories <- subjects |>
    rowwise() |>
    reframe(constant_trajectory(pick(everything()), fit$cuts))

trajectories
```

Now that we have trajectories over which we want predictions, we can predict the
survival probabilities over time.

```{r predict}
times <- seq(0, max(fit$cuts) * 0.95, length.out = 200)
S_mat <- predict_survival(fit, newdata = trajectories, times = times)
```

```{r plot}
as_tibble(S_mat, rownames = "study_id") |>
    pivot_longer(-study_id, names_to = "time", values_to = "S") |>
    mutate(
        time = as.numeric(time),
        age_at_start = subjects$age_at_start[match(study_id, subjects$study_id)]
    ) |>
    ggplot(aes(time, S, colour = factor(age_at_start), group = study_id)) +
    geom_line(linewidth = 0.8) +
    scale_colour_brewer(palette = "Set2", name = "Age at start") +
    scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
    labs(x = "Time (days)", y = "S(t | age_at_start)") +
    theme_minimal(base_size = 13)
```
